{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-11 thesis/landslide.tex
finish DPOR friendly example and rewrite sleep sets description to follow

In this section $t_i$ will denote a transition between two program states during execution,
with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
and $T(t_i)$ will denote the thread which was scheduled to produce that transition.

@@

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and $\bigstar$2)
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in $\bigstar$2,
the reordered {\tt tmp2 <- x} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting {\tt x <- tmp2}.

@@

However, as the legend spoils, this subset of the state space still contains redundant interleavings,
arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
Visual inspection shows that $\bigstar$5, while locally justified in trying to reorder {\tt tmp1 <- x} before {\tt x <- tmp2},
effectively serves % oops
only to reorder it with {\tt tmp2 <- x}
relative to the first symmetric subtree.
Landslide identifies this equivalence by checking,
%in this case during the iteration of DPOR after interleaving K,
in this case after K,
that if the first event to be reordered (here, {\tt tmp <- x})
has already been tested in an equivalent reordering around any number of preceding events (here, {\tt tmp2 <- x} and {\tt tmp2++}),
then the newly marked subtree is safe to prune.
The next section describes the implementation of this check in general.

@@

% TODO: condense this by removing the 'new' example and just continuing with the previous section's
In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
For example, consider the execution sequence
in Figure~\ref{fig:sleepsets}(a).
After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
Figure~\ref{fig:sleepsets}(b).
However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
so from this branch,
it next tags $I_{c'1}$, causing
Figure~\ref{fig:sleepsets}(c).
to be tested next, which is regrettably equivalent to the original interleaving
($t_a$ and $t_b$ having been reordered all told, which are independent).
To recover our forgotten original purpose,
which was already fulfilled by testing the second interleaving,
DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
for an already-explored interleaving beginning with the target thread.
If such exists,
%then because of the depth-first search ordering,
the new interleaving is guaranteed to be equivalent to one already checked,
and can be skipped.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{p{0.3\textwidth}p{0.25\textwidth}p{0.35\textwidth}}
	\begin{tabular}{l}
	$T_1$@$t_a$: \texttt{write~x} \\
	$T_2$@$t_b$: \texttt{read~y} \\
	$T_2$@$t_c$: \texttt{read~x} \\
	\end{tabular}
	&
	\begin{tabular}{l}
	$T_2$@$t_{b'}$: \texttt{read~y} \\
	$T_2$@$t_{c'}$: \texttt{read~x} \\
	$T_1$@$t_{a'}$: \texttt{write~x} \\
	\end{tabular}
	&
	\begin{tabular}{l}
	$T_2$@$t_{b'}$: \texttt{read~y} \\
	$T_1$@$t_{a''}$: \texttt{write~x} \\
	$T_2$@$t_{c''}$: \texttt{read~x} \\
	\end{tabular}
	\\
	\\
	(a) Original interleaving.
	&
	(b) Goal interleaving.
	&
	(c) Redundant interleaving.
	\end{tabular}
	\end{center}
	\caption{Motivating example for the sleep sets optimization.
	For clarity, threads are labeled here with numbers ($T_n$) and the transitions thereby with letters ($t_i$).}
	\label{fig:sleepsets}
\end{figure}

Landslide implements this check in {\tt equiv\_already\_explored()}.
This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.
Landslide's implementation differs from prior work by not explicitly tracking
sets of reordered threads and memory accesses they should expect to conflict with,
and instead identifies where the reduction should take place in subsequent iterations of DPOR.
% KEEP_RUNNING_DPORS_CHOSEN_TID
This approach also relies on the search ordering strategy ({\tt arbiter\_choose()})
to prefer scheduling the thread previously chosen for reordering by DPOR,
to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
Further optimizations such as {\em source sets} and {\em wakeup trees},
which achieve optimality (i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
are not yet implemented,
although to the best of my knowledge,
they provide further reduction only in cases of 3 or more threads;
I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.

% TODO: make sure it's clear that the quicksand chapter expts were run before this optmz was added,
% and that the HTM expts fro msigbovik paper were re-run for its chapater herein

>>>>

In this section $t_i$ will denote a transition between two program states during execution,
with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
and $T(t_i)$ will denote the thread which was scheduled to produce that transition.
A visual example will be given at the end to help reinforce the intuition behind the formalism.

@@

% trying to approximate the pastel background colors in something dark enough to render text in
\newcommand\dporTA[1]{\hilight{darklavender}{#1}\xspace}
\newcommand\dporTB[1]{\hilight{goldish}     {#1}\xspace}
\newcommand\dporTAcode[1]{\dporTA{\ensuremath{\mathbf{T_1}}: {\tt #1}}\xspace}
\newcommand\dporTBcode[1]{\dporTB{\ensuremath{\mathbf{T_2}}: {\tt #1}}\xspace}

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and $\bigstar$2)
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in $\bigstar$2,
the reordered \dporTB{{\tt tmp2 <- x}} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting \dporTB{{\tt x <- tmp2}}.

@@

@@

In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
As the presence of more equivalence arrows in Figure~\ref{fig:dpor-example-2} hints,
its reduced subset state space still contains redundancy,
arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
Visual inspection shows that $\bigstar$4, while locally justified
in trying to reorder \dporTA{{\tt tmp1 <- x}} before \dporTB{{\tt x <- tmp2}},
effectively serves % oops
only to reorder it with \dporTB{{\tt tmp2 <- x}}
relative to the first symmetric subtree ($\bigstar$1).
In other words,
even though DPOR marked each new branch with the intent only to reorder conflicting accesses,
$\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
Figure~\ref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{p{0.30\textwidth}p{0.30\textwidth}p{0.30\textwidth}}
	\begin{center}
	\begin{tabular}{l}
		\dporTAcode{tmp1 <- x} \\
		\dporTBcode{tmp2 <- x} \\
		\dporTAcode{x <- tmp1} \\
		\dporTBcode{x <- tmp2} \\
		% old example, pre dpor-example figure
		%$T_1$@$t_a$: \texttt{write~x} \\
		%$T_2$@$t_b$: \texttt{read~y} \\
		%$T_2$@$t_c$: \texttt{read~x} \\
	\end{tabular}
	\end{center}
	&
	\begin{center}
	\begin{tabular}{l}
		\dporTBcode{tmp2 <- x} \\
		\dporTBcode{x <- tmp2} \\
		\dporTAcode{tmp1 <- x} \\
		\dporTAcode{x <- tmp1} \\
		%$T_2$@$t_{b'}$: \texttt{read~y} \\
		%$T_2$@$t_{c'}$: \texttt{read~x} \\
		%$T_1$@$t_{a'}$: \texttt{write~x} \\
	\end{tabular}
	\end{center}
	&
	\begin{center}
	\begin{tabular}{l}
		\dporTBcode{tmp2 <- x} \\
		\dporTAcode{tmp1 <- x} \\
		\dporTAcode{x <- tmp1} \\
		\dporTBcode{x <- tmp2} \\
		%$T_2$@$t_{b'}$: \texttt{read~y} \\
		%$T_1$@$t_{a''}$: \texttt{write~x} \\
		%$T_2$@$t_{c''}$: \texttt{read~x} \\
	\end{tabular}
	\end{center}
	\\
	\begin{center}
	(a) Original branch (C).
	\end{center}
	&
	\begin{center}
	(b) Goal branch (K).
	\end{center}
	&
	\begin{center}
	(c) Redundant branch (L).
	\end{center}
	\end{tabular}
	\end{center}
	\caption{Motivating example for the sleep sets optimization,
	highlighting three interleavings from Figure~\ref{fig:dpor-example-2}.
	The always-independent {\tt tmpN++}s are omitted for brevity.
	}
	\label{fig:sleepsets}
\end{figure}

Intuitively speaking,
when DPOR entered the $\bigstar$3 subtree,
it did not ``remember'' which memory conflict it wanted to reorder \dporTA{$\mathbf{T_1}$} around
(i.e., that \dporTB{{\tt x <- tmp2}} should come before \dporTA{{\tt tmp1 <- x}}).
Upon witnessing the conflict in the new (intended) order,
it then tried to reorder it again,
producing interleavings regrettably equivalent to ones already tested
(all told, only \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}} having been reordered around \dporTA{{\tt tmp1 <- x}}).
%
To ``remember'' the original purpose of testing subtree $\bigstar$3,
which was already fulfilled by testing K,
DPOR can check just before tagging a new subtree (here, $\bigstar$4)
among all preceding transitions independent with the conflicting one
(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}} independent with \dporTA{{\tt tmp1 <- x}})
for an already-explored interleaving beginning with the target thread.
%(here, the original sibling of $\bigstar 3$).
If such exists,
the new subtree is guaranteed to be equivalent to one already checked,
and can safely be skipped.

Landslide implements this check in {\tt equiv\_already\_explored()},
which checks (in this case after executing K),
that if the first event to be reordered (here, \dporTA{{\tt tmp1 <- x}})
has already been tested in an equivalent reordering around any number of preceding events
(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}}),
then the newly marked subtree is safe to prune.
This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
so named because it effectively puts \dporTA{$\mathbf{T_1}$}
``to sleep'' until after the true conflicting access of \dporTB{{\tt x <- tmp2}}.

Landslide's implementation differs from prior work,
which explicitly tracks sets of reordered threads and expected conflicting accesses,
by instead identifying
where the reduction should occur
during subsequent DPOR iterations.
% KEEP_RUNNING_DPORS_CHOSEN_TID
This approach also relies on the search ordering strategy ({\tt arbiter\_choose()})
to prefer scheduling the thread previously chosen for reordering by DPOR,
to ensure the conflicting access happens before the preempted thread gets a chance to run again.
%to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
Further optimizations such as {\em source sets} and {\em wakeup trees},
which prior work has shown achieve optimality
(i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
are not yet implemented.
To the best of my knowledge,
they provide further reduction only in cases of 3 or more threads;
I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.

Chapter~\ref{chap:quicksand}'s experiments and Chapter~\ref{chap:education}'s user studies
were conducted before this optimization was implemented.
Note that its absence has no bearing on DPOR's soundness, only its efficiency,
and that Landslide showed good bug-finding performance even without it.
Chapter~\ref{chap:tm}'s experiments include this optimization,
because its presence was required to fairly compare the other reduction strategies presented therein.
